### 2019-04-24 papers/haskell19/3-representation.tex
Section 4

\begin{code}
data Fix (f :: * -> *) = Fix { unFix :: (f (Fix f)) }
\end{code}
%
Given a parametric type |f| of kind |* -> *| representing some target data type,
the type |Fix f| instantiates each recursive field of |f| with |Fix f|, thus
closing the definition of |f| into itself---note thus that the kind of |Fix f|
results |*|.

>>>>

\begin{code}
newtype Fix (f :: * -> *) = Fix { unFix :: f (Fix f) }
\end{code}
%
Given a parametric type |f| of kind |* -> *| representing some target data type,
the type |Fix f| instantiates each recursive field of |f| with |Fix f|, thus
closing the definition of |f| into itself---note thus that the kind of |Fix f|
results |*|.
%
In addition, note that we defined |Fix| as a |newtype|.
%
This way, the compiler will avoid producing a runtime representation for it,
thus avoiding unnecessary overhead.
%
This is possible since we only use it as a wrapper to obtain a self-recursive
type from a parametric one.

### 2019-04-24 papers/haskell19/4-generators.tex
Section 4

Even though we have extended QuickCheck's generators to be depth-aware, here we
also need to consider the parametric nature of our representations.
%
In the previous section, we defined each construction representation as being
parametric on the type of the recursive sub-terms, as a manner to defer this
choice until we have specified the overal shape of our target data.
%
Hence, each construction representation is of kind |* -> *|.
%
If we want to define our generators in a modular way, we also need to
parametrize somehow the generation of the recursive subterms!
%
Fortunately, QuickCheck already defines a lifted type class |Arbitrary1| for
types for parametric types of kind |*->*|:

@@

This type class allows us to write modular generators, leaving unspecified the
specific generator used to generate values of its type argument.
%
For instance, the default parametric generator for the data type |Maybe| looks
like:

@@

Then, we can use the same mechanism for our modular generators, extending it to
be depth-aware as follows:

\begin{code}
class BArbitrary1 (f :: * -> *) where
  liftBGen :: BGen a -> BGen (f a)
\end{code}

Note the similarity between the |Arbitrary1| and |BArbitrary1| type classes.
%
We will use this type class to implement random generators for each construction
we are considering.
%
Recalling our |Html| example, we can finally define modular random generator for
the constructions representing data constructors as follows:

@@

The rest of our construction representations can be generated using the same
idea.
%
For example, the |BArbitrary1| instances for |Fun_bold| and |Pat_simplify_2| are
as follows:

@@

The previous instance uses the |BArbitrary1| of our final representation to
generate sub-terms recursively using itself (|bounded d|) as the parametrized
input generator, wrapping each recursive sub-term with a |Fix| data constructor.

The machinery developed so far lets us generate single random constructions in a
modular fashion.
%
However, we still need to develop our generation mechanism a bit further in
order to generate composite representations built using the |oplus''| type-level
operator.
%
This is the objective of the next sub-section.

% --------------------------------------
% Encoding Generation Instructions Using Types
%
\subsection{Encoding Generation Instructions Using Types}

\begin{code}
newtype ((f :: * -> *) otimes (n :: Nat)) a = Freq (f a)
\end{code}

@@

\begin{code}
instance Algebra f a => Algebra (f otimes n) a where
  alg (Freq f) = alg f
\end{code}

\begin{code}
instance Algebra f a => Algebra (Term f) a where
  alg (Term f) = alg f
\end{code}

\begin{code}
instance BArbitrary1 f => BArbitrary1 (f otimes n) where
  liftBGen gen d = Freq <$> liftBGen gen d

instance BArbitrary1 f => BArbitrary1 (Term f) where
  liftBGen gen d = Term <$> liftBGen gen d
\end{code}

\begin{code}
type family FreqOf (f :: * -> *) :: Nat where
  FreqOf (f oplus g)   = FreqOf f + FreqOf g
  FreqOf (f otimes n)  = n * FreqOf f
  FreqOf (Term f)      = FreqOf f
  FreqOf _             = 1
\end{code}

\begin{code}
type family FreqOf' (f :: * -> *) :: Nat where
  FreqOf' (f oplus g)   = FreqOf' f  +  FreqOf' g
  FreqOf' (f otimes n)  = n  *  FreqOf' f
  FreqOf' (Term f)      = FreqOf f
  FreqOf' _             = 0
\end{code}

@@

@@

\begin{code}
genFix  ::  forall f dot BArbitrary1 f =>  BGen (Fix f)
genFix d = bounded at(Fix f) d
\end{code}

\begin{code}
genEval :: forall f a dot (BArbitrary1 f, Algebra f a) => BGen a
genEval d = eval <$> genFix at f d
\end{code} %$

>>>>

Even though we have extended QuickCheck's generators to be depth-aware, here we
also need to consider the parametric nature of our representations.
%
In the previous section, we defined each construction representation as being
parametric on the type of its recursive sub-terms, as a way to defer this choice
until we have specified the overal shape of our target data.
%
Hence, each construction representation is of kind |* -> *|.
%
If we want to define our generators in a modular way, we also need to
parametrize somehow the generation of the recursive subterms!
%
If we look at QuickCheck, this library already defines a type class |Arbitrary1|
for parametric types of kind |* -> *|:

@@

This type class allows us to write modular generators, leaving unspecified the
specific generator used when generating values of its type argument.
%
For instance, the default parametric generator for the data type |Maybe| looks
like:

@@

Where the random generator use to generate the value stored in a |Just|
constructor is parametrized by the |gen| input variable.
%
We can use this same mechanism for our modular generators, first extending it to
be depth-aware as follows:

\begin{code}
class BArbitrary1 (f :: * -> *) where
  liftBGen :: BGen a -> BGen (f a)
\end{code}
%
Note the similarity between the |Arbitrary1| and |BArbitrary1| type classes.
%
Then, we will use this type class to implement random generators for each
construction we are considering.
%
Recalling our |Html| example, we can define modular random generator for the
constructions representing data constructors as follows:

@@

The rest of our construction representations can be generated analogously.
%
For example, the |BArbitrary1| instances for |Fun_bold| and |Pat_simplify_2| are
as follows:

@@

The previous instance employs the |BArbitrary1| of our final representation to
generate sub-terms recursively using itself (|bounded d|) as the parametrized
input generator, wrapping each recursive sub-term with a |Fix| data constructor.

%
% Wrapping this subsection
%
The machinery developed so far lets us generate single random constructions in a
modular fashion.
%
However, we still need to develop our generation mechanism a bit further in
order to generate composite representations built using the |oplus| type-level
operator.
%
This is the objective of the next sub-section.

% --------------------------------------
% Encoding Generation Behavior Using Types
%
\subsection{Encoding Generation Behavior Using Types}

As we have shown so far, generating each construction representation is rather
straightforward: there is only one data constructor to select, and every field
is generating following a mechanical recipe.
%
In our approach, most of the generation complexity is encoded in the random
generator used for composite representations built upon the |oplus| operator.
%
Before introducing it, we need to define some additional machinery to encode the
notions of terminal construction and generation frequency.

Recalling the random generator for |Html| presented in Section
\ref{sec:sources}, we can observe that the last generation level (see |gen 0|)
is constrained to generate only the non-recursive subset of all the available
constructions.
%
In order to model this behavior in our framework, we can start by defining a
data type |Term| to ``tag'' every terminal construction explicitly:

@@

Then, if |f| is a non-recursive construction, the type |Term f oplus' g| can be
interpreted as representing values generated using values drawn both from |f|
and |g|, but ``closed'' using only values from |f|.
%
Since this data type will not add any semantic information to our represented
values, we can define suitable |Algebra| and |BArbitrary1| instances for it
simply by delegating the work to the inner type:

\begin{code}
instance Algebra f a => Algebra (Term f) a where
  alg (Term f) = alg f

instance BArbitrary1 f => BArbitrary1 (Term f) where
  liftBGen gen d = Term <$> liftBGen gen d
\end{code} %$

The next building block of our framework consists in a way of specifying the
generation frequency of each construction.
%
For this purpose, we can follow the same reasoning as before, defining a data
type-level operator |otimes| to explicitly tag the generation frequency of a
given representation:

\begin{code}
newtype ((f :: * -> *) otimes'' (n :: Nat)) a = Freq (f a)
\end{code}

In the definition above, the operator |otimes| is parametrized by a type-level
natural number |n| (of kind |Nat|) representing the desired generation
frequency.
%
In this light, the type |(f otimes'' 3) oplus'' (g otimes'' 1)| represents
values generated using values from both |f| and |g|, where |f| is randomly
chosen three times more frequently than |g|.
%
In practice, we defined |otimes| such that it associates more strongly than
|oplus|, thus avoiding the need of parenthesis in types like the previous one.
%
Analogously as |Term|, the operator |otimes| does not add any semantic
information to the values it represents, so we can define its |Algebra| and
|BAbitrary1| instance by delegating the work to the inner type as before:

\begin{code}
instance Algebra f a => Algebra (f otimes'' n) a where
  alg (Freq f) = alg f

instance BArbitrary1 f => BArbitrary1 (f otimes'' n) where
  liftBGen gen d = Freq <$> liftBGen gen d
\end{code} %$

With these two new type level combinators |Term| and |otimes|, we are now able
to express the behavior of our entire generation process based solely on the
type we are generating.
%
In addition to these combinators, we will need to perform some type-level
computations based on them in order to define our random generator for composite
representations.
%
Consider for instance the following type defined using parenthesis for clarity:

\begin{code}
(f otimes'' 2) oplus'' ((g otimes'' 3) oplus'' (Term h))
\end{code}

Our generation process will traverse this type one combinator at a time,
processing each occurrence of |oplus| independently.
%
This means that, in order to select the appropriate generation frequency of each
operand we need to calculate the overall sum of frequencies on each side of the
|oplus|.
%
For this purpose, we can implement a closed type family |FreqOf| to compute the
overal sum of frequencies of a given representation type:

\begin{code}
type family FreqOf (f :: * -> *) :: Nat where
  FreqOf (f oplus'' g)   = FreqOf f + FreqOf g
  FreqOf (f otimes'' n)  = n * FreqOf f
  FreqOf (Term f)        = FreqOf f
  FreqOf _               = 1
\end{code}

This type-level function takes a representation type as an input and traverses
it recursively, adding up each frequency tag found in the process, and returning
a type-level natural number.
%
Note how in the second equation we multiply the frequency encoded in the
|otimes| tag with the frequency of the type it is wrapping.
%
This way, the type |((f otimes'' 2) oplus'' g) otimes'' 3| is equivalent to |(f
otimes'' 6) oplus'' (g otimes'' 3)|, following the natural intuition for the
addition and multiplication operations over numbers.
%
Moreover, if a type does have an explicit frequency then its frequency is simply
defaulted to one.

On the other hand, the last step of our generation process which only generates
terminal constructions could be seen as generating values considering all the
available constructions, but instead treating the non-terminal ones as having
generation frequency zero.
%
This way we can reuse our previous type level computation to calculate the
\emph{terminal generation frequency} |FreqOf'| of a given representation type as
follows:

\begin{code}
type family FreqOf' (f :: * -> *) :: Nat where
  FreqOf' (f oplus'' g)   = FreqOf' f  +  FreqOf' g
  FreqOf' (f otimes'' n)  = n  *  FreqOf' f
  FreqOf' (Term f)        = FreqOf f
  FreqOf' _               = 0
\end{code}

Similar to |FreqOf|, the type family above traverses its input type adding the
terminal frequency of each sub-type.
%
However, |FreqOf'| only considers the frequency of those representation
sub-types that are explicitly tagged as terminal, returning zero in any other
case.

Using the typle level combinators introduced at the beginning of this
sub-section along with the previous type-level computations over frequencies, we
are finally in position of defining our random generator for composite
representations:

@@

Like the generator introduced in Section \ref{sec:sources} for |Html|, this
generator branches over the current generation depth |d|.
%
In the case we can still generate values from any construction (|d > 0|), we
will use QuickCheck's |frequency| operation to randomly choose between
generating a value of each side of the |oplus|, i.e., either a value of |f| or a
value of |g|, following the generation frequencies specified for both of them,
and wrapping the values with the appropriate injection |InL| or |InR| on each
case.
%
Such frequencies are obtained by \emph{reflecting} the type-level natural values
obtained from applying |FreqOf| to both |f| and |g|, using a type-dependent
function |freqVal| that returns the number corresponding to the type-level
natural value we apply to it:

@@

Note that the type of |freqVal| is ambiguous, since it quantifies over every
possible known type-level natural value |n|.
%
We need to use a \emph{visible type application} (employing the |at (...)|
syntax) to desambiguate which type-level natural value we are refering to.
%
Then, for instance, the value |freqVal at (FreqOf (f otimes'' 5))| will evaluate
to |5 :: Int|.

The remaining part of our random generator works analogously, except that this
time we only want to genrate terminal constructors, hence we use the |FreqOf'|
type family to compute the terminal generation frequency of each operand.
%
If any of |FreqOf' f| or |FreqOf' g| evaluates to zero, then it means that that
operand does not contain any terminal construction, and |frequency| will not
consider it when generating terminal values.

If it happens that both |FreqOf' f| and |FreqOf' g| compute to zero when
genrating a composite value, then it will produce a runtime error triggered by
the function |frequency|, as it does not have anything with a positive frequency
to generate.
%
This kind of exceptions will arise if we forget to tag at least a single
construction within a composite representation as terminal.
%
Fortunately, such runtime exceptions can be catched in compile time by defining
a type constraint that ensures we are trying to generate values using a
representation with a strictly positive terminal generation frequency---thus
containing at least a single terminal construction:

\begin{code}
type family HasTerminals (f :: * -> *) :: Constraint where
  HasTerminals f = IsPositive (FreqOf' f)

type family IsPositive (n :: Nat) :: Constraint where
  IsPositive 0 = TypeError "No terminals"
  IsPositive _ = ()
\end{code}
%
The previous type families compute the terminal generation frequency of a
representation type |f|, returning a type error if its result is zero, or an
empty constraint |()| that is always trivially satisfied in any other case.
%
Finally, we can use this constraint to define a safe generation primitive
|genEval| to obtain a random depth-bounded generator for every target type |a|
specified using a safe generation representation |f|:

\begin{code}
genEval  :: forall f a dot
  (BArbitrary1 f, HasTerminals f, Algebra f a) => BGen a
genEval d = eval <$> bounded at (Fix f) d
\end{code} %$

Note how this primitive is also ambiguous in the type used for the
representation.
%
This allows us to use a visible type application to obtain values from the same
target type but generated using a different underlying representations.

So far we have seen how to represent and generate values for a target data type
by combining a wide selection of structural constructions, as well as a series
of type-level combinators to encode the generation behavior.
%
The next section refines our type-level machinery in order to export an elegant
domain specific language for specifying composable random generators.
